---
title: "Project"
author: "Manasa"
date: "2024-03-17"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,echo = TRUE)
```

## R Markdown

```{r cars}
if (!requireNamespace("readr", quietly = TRUE)) install.packages("readr")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("tidyr", quietly = TRUE)) install.packages("tidyr")
if (!requireNamespace("lubridate", quietly = TRUE)) install.packages("lubridate")

#Load the libraries
library(lubridate)
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)

crd <- read_csv("/Users/manasavishnumoorthy/Desktop/IDMP/CRD.csv", 
                 na = c("", "NA", " "), show_col_types = FALSE)

incident <- read_csv("/Users/manasavishnumoorthy/Desktop/IDMP/incidents_data.csv", 
                 na = c("", "NA", " "), show_col_types = FALSE)

non_motorists <- read_csv("/Users/manasavishnumoorthy/Desktop/IDMP/non_motorist_data.csv", 
                 na = c("", "NA", " "), show_col_types = FALSE)

#Convert 'Local Case Number' to character in all data frames
crd$`Local Case Number` <- as.character(crd$`Local Case Number`)
non_motorists$`Local Case Number` <- as.character(non_motorists$`Local Case Number`)
incident$`Local Case Number` <- as.character(incident$`Local Case Number`)

combined_data <- crd %>%
  full_join(non_motorists, by = c("Report Number", "Local Case Number")) %>%
  full_join(incident, by = c("Report Number", "Local Case Number"))

#Check the first few rows of the combined data
#str(combined_data)

```

**Data preprocessing**

**Standardize text variables:**
Convert all character variables to lowercase for consistency.
```{r}
combined_data <- combined_data %>%
  mutate(across(where(is.character), tolower))
```

**Handle missing values:**
Replace missing values in character columns with "unknown" and in numeric columns with 0 (or another appropriate value).
```{r}
combined_data <- combined_data %>%
  mutate(across(where(is.character), ~replace_na(., "unknown"))) %>%
  mutate(across(where(is.numeric), ~replace_na(., 0)))

```

**Create new variables from 'Crash Date/Time':**
Parse the 'Crash Date/Time' and extract useful components such as the date, hour, day of the week, etc.
```{r}

combined_data <- combined_data %>%
  mutate(`Crash Date/Time` = mdy_hms(`Crash Date/Time.x`))  # Convert to date-time object

# Extract components
combined_data <- combined_data %>%
  mutate(
    Date = as.Date(`Crash Date/Time`),  # Extract date
    Hour = hour(`Crash Date/Time`),  # Extract hour
    Weekday = wday(`Crash Date/Time`, label = TRUE),  # Extract day of the week
    Month = month(`Crash Date/Time`, label = TRUE)  # Extract month
  )
```

```{r}
#str(combined_data)
```


**Substance Abuse Factor:**
Investigating the factor of substance abuse in collisions.
```{r}
combined_data$`Substance Abuse` <- ifelse(combined_data$`Driver Substance Abuse.x` != "none detected" & 
                                          combined_data$`Driver Substance Abuse.x` != "unknown", "Yes", "No")

ggplot(combined_data, aes(x = `Substance Abuse`, fill = `Substance Abuse`)) +
  geom_bar() +
  labs(title = "Substance Abuse Factor in Collisions", x = "Substance Abuse Involved", y = "Count")
```

**Accidents by Light Conditions and Time of Day:**
It might be interesting to see how light conditions impact accidents, and if there are peak times.
```{r}
library(ggplot2)
library(dplyr)

#Filter out unwanted categories
filtered_data <- combined_data %>%
  filter(!Light.x %in% c("n/a", "other", "unknown"))

# Count the number of accidents by light condition
light_data <- filtered_data %>%
  count(Light.x) %>%
  rename(Number_of_Accidents = n)

# Create a pie chart
ggplot(light_data, aes(x = "", y = Number_of_Accidents, fill = Light.x)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +  # This makes it a pie chart
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Proportion of Accidents by Light Conditions", x = "", y = "", fill = "Light Condition") +
  theme_void() 

```
**Heatmap of Collisions by Hour and Day of the Week**
This visualization will help us understand when most collisions occur during the week.
```{r}
library(ggplot2)
library(dplyr)

# Create a column for the day of the week and hour
combined_data$DayOfWeek <- weekdays(as.Date(combined_data$`Crash Date/Time`))
combined_data$Hour <- format(combined_data$`Crash Date/Time`, "%H")

# Prepare the data for the heatmap
heatmap_data <- combined_data %>%
  filter(!is.na(DayOfWeek) & DayOfWeek != "na") %>%  # Filter out the "NA" or "na" entries
  group_by(DayOfWeek, Hour) %>%
  summarize(Count = n(), .groups = 'drop') %>%
  mutate(DayOfWeek = factor(DayOfWeek, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")))

# Create the heatmap
ggplot(heatmap_data, aes(x = Hour, y = DayOfWeek, fill = Count)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Heatmap of Collisions by Hour and Day of Week", x = "Hour of Day", y = "Day of Week") +
  theme_minimal()

```



```{r}
library(ggplot2)
library(dplyr)


collisions_by_day_filtered <- combined_data %>%
  filter(!is.na(DayOfWeek) & DayOfWeek != "na") %>%  # Filter out the "NA" or "na" entries
  group_by(DayOfWeek) %>%
  summarise(NumberOfCollisions = n(), .groups = 'drop') %>%
  mutate(DayOfWeek = factor(DayOfWeek, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")))


ggplot(collisions_by_day_filtered, aes(x = DayOfWeek, y = NumberOfCollisions, fill = DayOfWeek)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = NumberOfCollisions), vjust = -0.3) +
  labs(title = "Number of Collisions by Day of the Week", x = "Day of the Week", y = "Number of Collisions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

**ARIMA**


```{r}
library(dplyr)
library(lubridate)
library(forecast)
library(tibble)
library(ggplot2)


# Convert the Crash Date/Time to a proper Date format 
combined_data$Crash_Date <- as.Date(mdy_hms(combined_data$`Crash Date/Time.x`))

# Remove rows with NA in Crash_Date 
combined_data <- combined_data %>% 
  filter(!is.na(Crash_Date))

# Aggregate data by Crash_Date to get the number of collisions per day
daily_collisions <- combined_data %>% 
  group_by(Crash_Date) %>% 
  summarise(Number_of_Collisions = n(), .groups = 'drop')

# Generate the time series object from daily collision counts
collision_ts <- ts(daily_collisions$Number_of_Collisions, frequency = 365)

# Fit the ARIMA model
fit <- auto.arima(collision_ts)

# Check model summary
summary(fit)
```


```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Ensure that the dates are in the proper Date format
daily_collisions$Crash_Date <- as.Date(daily_collisions$Crash_Date)

#Plot using ggplot2
#ggplot(daily_collisions, aes(x = Crash_Date, y = Number_of_Collisions)) +
# geom_line(color = "blue") +
#  geom_smooth(method = "lm", color = "red", se = FALSE) + # Trend line
#  labs(title = "Historical Collision Data",
#       x = "Date",
#       y = "Number of Collisions") +
#  theme_minimal()
#

ggplot(daily_collisions, aes(x = Crash_Date, y = Number_of_Collisions)) +
  geom_line(color = "blue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3), color = "red", se = FALSE) + # Quadratic trend line
  labs(title = "Historical Collision Data with Non-linear Trend",
       x = "Date",
       y = "Number of Collisions") +
  theme_minimal()
```

```{r}
library(forecast)
library(ggplot2)

# Forecast the next 30 days using the fitted ARIMA model
forecast_future <- forecast(fit, h = 30)

# Create a data frame for the forecast dates and predictions
last_date <- max(daily_collisions$Crash_Date)  # Last date in your historical data
forecast_dates <- seq.Date(from = last_date + 1, by = "day", length.out = 30)

forecast_data <- data.frame(
  Date = forecast_dates,
  Forecast = as.numeric(forecast_future$mean),
  Lower_80 = forecast_future$lower[, "80%"],
  Upper_80 = forecast_future$upper[, "80%"],
  Lower_95 = forecast_future$lower[, "95%"],
  Upper_95 = forecast_future$upper[, "95%"]
)


# Plot the forecast using ggplot2
ggplot(forecast_data, aes(x = Date)) +
  geom_line(aes(y = Forecast), color = "red") +
  geom_ribbon(aes(ymin = Lower_95, ymax = Upper_95), fill = "grey80", alpha = 0.2) +
    geom_ribbon(aes(ymin = Lower_80, ymax = Upper_80), fill = "grey60", alpha = 0.4) +
  labs(title = "30 Day Collision Forecast",
       x = "Date",
       y = "Forecasted Number of Collisions") +
  theme_minimal()
```

```{r}

# Time series decomposition
decomposition <- stl(collision_ts, s.window = "periodic")

# Plot decomposed components
plot(decomposition)
```

```{r}
# Model selection
# Example: Fit a SARIMA model
sarima_model <- auto.arima(collision_ts, seasonal = TRUE)

# Model validation
# Diagnostic plots
plot(sarima_model)
```

```{r}
# Forecasting
forecast_values <- forecast(sarima_model, h = 30)

# Evaluate forecast accuracy
accuracy(forecast_values)
```

**Regression Analysis**
```{r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(caret)

# Assuming 'Injury Severity.x' is a factor variable indicating the severity of crashes
# Let's encode it into numeric values for regression analysis
combined_data$Injury_Severity_Num <- as.numeric(factor(combined_data$`Injury Severity.x`))

# Selecting different variables for the regression model
model_data <- combined_data %>%
  select(`Injury_Severity_Num`, `Weather`, `Surface Condition`, `Light`, `Vehicle Year`, `Vehicle Make`, `Vehicle Model`)

# Convert Weather, Surface_Condition, Light, Vehicle_Make, and Vehicle_Model to factors
model_data$Weather <- as.factor(model_data$Weather)
model_data$Surface_Condition <- as.factor(model_data$`Surface Condition`)
model_data$Light <- as.factor(model_data$Light)
model_data$Vehicle_Year <- as.factor(model_data$`Vehicle Year`)
model_data$Vehicle_Make <- as.factor(model_data$`Vehicle Make`)
model_data$Vehicle_Model <- as.factor(model_data$`Vehicle Model`)

# Encoding factors into numeric values for regression analysis
model_data <- model_data %>%
  mutate(Weather_Num = as.numeric(Weather),
         Surface_Condition_Num = as.numeric(Surface_Condition),
         Light_Num = as.numeric(Light),
         Vehicle_Make_Num = as.numeric(Vehicle_Make),
         Vehicle_Model_Num = as.numeric(Vehicle_Model))

# Perform multiple linear regression
model <- lm(Injury_Severity_Num ~ Weather_Num + Surface_Condition_Num + Light_Num + Vehicle_Year + Vehicle_Make_Num + Vehicle_Model_Num, data = model_data)

# Summary of the regression model
summary(model)

# Predictions using the model
predictions <- predict(model)

# Visualize the actual vs. predicted values
ggplot(model_data, aes(x = predictions, y = Injury_Severity_Num)) +
  geom_point() +
  geom_smooth() +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
  labs(x = "Predicted Injury Severity", y = "Actual Injury Severity", title = "Actual vs. Predicted Injury Severity")
```



